---
title: "Blackjack simulations"
author: "Rob Kravec"
date: "December 1, 2020"
output: 
  html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(cumstats)
library(knitr)
library(parallel)
set.seed(1995)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,
                      message = FALSE, warning = FALSE, comment = NA,
                      fig.width=8, fig.height=6)
```

We'll approach this problem by writing several small functions that will be 
combined to play blackjack.

Initialize a shoe, which we won't alter in the course of playing blackjack

```{r create-shoe}
# Create initial deck. J, Q, K are all depicted as 10. Ace is depicted as 1
# but will sometimes be counted as 11 (see code for hit_me() function)
shoe <- rep(c(2:10, 10, 10, 10, 1), times = 4 * 7)
```

Create a function that returns a shuffled shoe with two 8's and one 9 removed

```{r new-shoe-function}
new_shoe <- function(shoe){
  # Remove the starting cards
  deck <- shoe[-c(which(shoe == 8)[1:2], which(shoe == 9)[1])]
  
  # Return shuffled deck
  return(sample(x = deck, size = length(deck), replace = F))
}
```

Define function for a person who decides to hit

```{r hit-me-function}
hit_me <- function(current_deck, personId) {
  
  # Show what hand value would be for each potential "hit"
  potential_hits <- cumsum(c(personId, current_deck))
  
  # Detect aces. Add 1 to account for extra element in potential_hits
  aces <- which(current_deck == 1) + 1
  
  # Check if there is an ace that needs to be incremented to an 11
  if(potential_hits[aces[1]] <= 11) {
    potential_hits[aces[1]: length(potential_hits)] <- 
      potential_hits[aces[1]: length(potential_hits)] + 10
  }
  
  # Find first instance greater than or equal to 17
  stop_index <- which(potential_hits >= 17)[1]
  
  # Return list that contains the score corresponding to the personId and the
  # remaining deck
  return(list(personId = potential_hits[stop_index],
         new_deck = current_deck[-c(1:stop_index)])
         )
}
```

Define function for a single hand using the "stand" strategy

```{r stand-function}
stand <- function() {
  # Designate starting values for dealer and player
  player <- 8 + 8
  dealer <- 9
  
  # Generate a new shuffled deck
  deck <- new_shoe(shoe = shoe)

  # Have dealer hit until achieving 17 or more
  dealer <- hit_me(current_deck = deck, personId = dealer)$personId
  
  # Check for dealer bust
  if (dealer > 21) {
    return(1)
  }
  
  # Return result
  return(player - dealer)
}
```

Define function for a single hand using the "hit" strategy

```{r hit-function}
hit <- function(player_start) {
  # Designate starting values for dealer and player
  player <- player_start
  dealer <- 9
  
  # Generate a new shuffled deck
  deck <- new_shoe(shoe = shoe)
  
  # Have player hit until achieving 17 or more
  player_hit <- hit_me(current_deck = deck, personId = player)
  
  # Early exit if player busts
  if(player_hit$personId > 21) {
    return(-1)
  }
  
  # Have dealer hit until achieving 17 or more
  dealer <- hit_me(current_deck = player_hit$new_deck, 
                   personId = dealer)$personId
  
  # Early exit if dealer busts
  if(dealer > 21) {
    return(1)
  }
  
    # Return result
  return(player_hit$personId - dealer)
}

```

Define function for a single hand using the "split" strategy

```{r split-function}
split <- function() {
  # Initialize vector of final player scores
  final_p_scores <- c()
  
  # Designate starting values for dealer and player
  player <- 8
  dealer <- 9
  
  # Generate a new shuffled deck
  deck <- new_shoe(shoe = shoe)
  
  # Create a counter denoting the number of player hands (due to splitting)
  hands <- 2
  
  # Perform as many splits as are needed
  while(hands > 0) {
    
    # Check to see whether another split will be needed
    if(deck[1] == 8) {
      hands <- hands + 1
    }
    
    # Play a round on a single hand
    player_hit <- hit_me(current_deck = deck, personId = player)
    hands <- hands - 1
    final_p_scores <- c(final_p_scores, player_hit$personId)
    deck <- player_hit$new_deck
  }
  
  # Have dealer hit until achieving 17 or more
  dealer <- hit_me(current_deck = player_hit$new_deck, 
                   personId = dealer)$personId
  
  # Early exit if dealer busts
  if(dealer > 21) {
    if(all(final_p_scores > 21)) {
      return(0) # We'll say it's a push if everyone busts
    }
    return(1)
  }
  
  # Recode hands in which the player busts to be an automatic defeat
  bust_hands <- which(final_p_scores > 21)
  final_p_scores[bust_hands] <- 1 # Use small value to create automatic defeat
  
  # Return results
  results <- final_p_scores - dealer
  wins <- sum(results > 0)
  losses <- sum(results < 0)
  if(wins > losses) {
    return(1)
  } else if (losses > wins) {
    return(-1)
  } else {
    return(0)
  }
}
```

Run 20,000 simulations of each function, using parallel processing, and display 
the results

```{r sim-parallel}
# Initiate parallel processes
x <- list()
nreps <- 20000
x$stand <- mcparallel(replicate(n = nreps, stand()))
x$hit <- mcparallel(replicate(n = nreps, hit(16)))
x$split <- mcparallel(replicate(n = nreps, split()))

# Collect results
probs <- mccollect(x)

# Calculate outcomes
outcomes <- mclapply(probs, 
                     function(x) c(mean(x > 0), mean(x < 0), mean(x == 0)),
                     mc.cores = 4)

# Display results
results_df <- data.frame(Outcome = c("Win", "Lose", "Push"),
                         Strategy = map(outcomes, `[`))
colnames(results_df) <- c("Outcome","Stand", "Hit", "Split")
kable(results_df)
```

Under these rules and without financial constraints, it would make sense to
"split."